**Covid-19 vaccination in the province of Ontario: A geographical and socio-economical analysis**

### 2023-02-21 data_cleaning/Data_cleaning_Feb_17.qmd
re-grouped data and ran model, fixed code

```{r, setup, echo=FALSE}
#| warning: false
library(tidyverse)
library(here)
library(lubridate)
library(patchwork)
library(visdat)
library(gt)
library(scales)
library(scico)

@@

t1<-d1%>%
  select(age_group,
         employed,
         h_size,
         income,race,
         remote_work,
         sick_leave,
         school,
         pc_1,
         pc_2,
         pc_3)%>%
 summarise_all(funs(sum(!is.na(.)/length(.))))%>%#count percentage of NAs
  pivot_longer(everything())%>%
  arrange(desc(value))%>%
  rename(observations=value,
         variable=name)

@@

clean_data<-d1%>%
  select(age_group,
         employed,
         age,
         income,
         race,
         city,
         h_size,
         date,
         first_dose,
         second_dose,
         get_vaccine)%>%
  drop_na(employed)

>>>>

```{r, setup, echo=FALSE}
#| warning: false
library(tidyverse)
library(here)
library(lubridate)
library(gtsummary)
library(patchwork)
library(visdat)
library(gt)
library(scales)
library(scico)

@@

t1<-d1%>%
  select(age_group,
         income,
         race,
         remote_work,
         sick_leave,
         school,
         pc_1,
         pc_2,
         pc_3)%>%
 summarise_all(funs(sum(!is.na(.)/length(.))))%>%#count percentage of NAs
  pivot_longer(everything())%>%
  arrange(desc(value))%>%
  rename(observations=value,
         variable=name)

@@

clean_data<-d1%>%
  select(age_group,
         income,
         race,
         city,
         first_dose)%>%
  drop_na(age_group,
          income,
          race,
          city,
          first_dose)

data_summary<-clean_data %>%
  tbl_summary(percent="row",
              by= first_dose,
              include=c(income,
                        age_group,
                        race)
  ) %>% bold_labels()

## original clean dataset

data1<-read.csv(here("data","clean_dataset.csv"))

data_summary2<-data1 %>%
  drop_na(race)%>%
  tbl_summary(percent="row",
              by= first_dose,
              include=c(income,
                        age_group,
                        race)
  ) %>% bold_labels()

tbl_merge(tbls=list(data_summary2,data_summary),tab_spanner = c("Original Clean Dataset","Dataset choosing covariates first"))

### 2023-02-21 manuscript/02-methods.qmd
re-grouped data and ran model, fixed code

Additionally, users who indicated they were under the age of 16 were exited from the survey, no record was created, and these users were unable to navigate back to the "age select" screen. The personal identifier information from each user was automatically scrubbed and replaced by a unique ID. Respondents were drawn exclusively from the province of Ontario, as per their devices meta-data.

## Survey responses

### Socio-demographic factors {#sec-socio-demographic-factors}

Socio-economic information selected for analysis from the survey answers included age group, income bracket, and race/ethnicity. Information about these socio-economic factors in the survey is provided in @tbl-covariates.

| Variable             | Values                                                                                                                          |
|:------------------|-----------------------------------------------------|
| Age group            | 16-24,25-34,35-44,45-54,55-64, 65+                                                                                              |
| Income bracket (CAD) | \<15,000, 15,000-24,999, 25,000-39,999, 40,000-59,999, 60,000-89,999, \>90,000                                                  |
| Race/ethnicity       | Arab/Middle Eastern, Black, East Asian/Pacific Islander, Indigenous, Latin American, Mixed, South Asian, White Caucasian, Other |
|                      |                                                                                                                                 |

@@

We used a fixed-effects logistic regression model to estimate the probability of vaccination depending on the socio-economic factors described in @sec-socio-demographic-factors and the Health Regions from @sec-geographical-location, as previous studies have shown that socio-economic factors and their interactions are significant predictors of intent of vaccination and vaccination status [@nguyen2022; @shih2021; @cnat2022a]. The model appears in @eq-model3,

$$
\begin{aligned}
\log \left( \frac{p\textrm{(vac)}}{1-p\textrm{(vac)}} \right) = \beta_0+ \beta_{1}\textrm{(Age group)} +\beta_{2} \textrm{ Race} + \beta_3 \textrm{ Health Region} + \beta_4 \textrm{ Income}+\\ \beta_5\textrm{(Health Region} \times \textrm{Income)} + \beta_6 \textrm{ (Age group} \times \textrm{Income)} + \beta_7 \textrm{ (Race} \times \textrm{Income)}
\end{aligned}
$$ {#eq-model3}

Where $p\textrm{(vac)}$ indicates the probability of having received the first dose of a Covid-19 vaccine, $\beta_0$ indicates the population intercept, $\beta_1...\beta_7$ indicate the coefficients for each of the fixed and interaction effects of age group, race, Health Region, and income. The the model was fitted using the function `svyglm` from the `survey` R package in order to incorporate the correction in sampling probability obtained from raking.

>>>>

Additionally, users who indicated they were under the age of 16 were exited from the survey without creating a record, furthermore, these users were unable to navigate back to the "age select" screen. The personal identifier information from each user that participated in the survey was automatically scrubbed and replaced by a unique ID. Respondents were drawn exclusively from the province of Ontario, as per their devices meta-data.

## Survey responses {#sec-socio-demographic-factors}

Socio-economic information selected from the survey answers included age group, income bracket, and race/ethnicity. The levels of each socio-economic factor used for analysis appear in @tbl-covariates.

| Variable             | Levels                                            |            
|:---------------------|---------------------------------------------------|
| Age group            | 16-34,35-54,55 and over                           |
| Income bracket (CAD) | under 25,000, 25,000-59,999, 60,000 and above     |
| Race/ethnicity       | Arab/Middle Eastern, Black, East Asian/Pacific Islander, Indigenous, Latin American, Mixed, South Asian, White Caucasian, Other |

@@

We used a logistic regression model to estimate the probability of vaccination depending on the socio-economic factors described in @sec-socio-demographic-factors, the Health Regions from @sec-geographical-location, and the interactions between Race and Health Region, and Race and income as previous studies have shown that socio-economic factors and their interactions are significant predictors of intent of vaccination and vaccination status [@nguyen2022; @shih2021; @cnat2022a]. The model appears in @eq-model3,

$$
\begin{aligned}
\log \left( \frac{p\textrm{(vac)}}{1-p\textrm{(vac)}} \right) = \beta_0+ \beta_{1}\textrm{(Age group)} +\beta_{2} \textrm{ Race} + \beta_3 \textrm{ Health Region} + \beta_4 \textrm{ Income}+\\ \beta_5\textrm{(Health Region} \times \textrm{Race)} + \beta_6 \textrm{ (Income} \times \textrm{Race)}
\end{aligned}
$$ {#eq-model3}

Where $p\textrm{(vac)}$ indicates the probability of having received the first dose of a Covid-19 vaccine, $\beta_0$ indicates the population intercept, and $\beta_1...\beta_6$ indicate the coefficients for each term. The the model was fitted using the function `svyglm` from the `survey` R package in order to incorporate the correction in sampling probability obtained from raking.

### 2023-02-21 manuscript/03-results.qmd
re-grouped data and ran model, fixed code

@tbl-model shows the results of the logistic regression on vaccination status using socio-economic, geographical factors (by Health Region) and their interactions as predictors. The reference groups in each case were set as follows: 16 to 24 years (age group), White Caucasian (Race), Toronto (Health Region), Over CAD 90,000 (Income). Socio-economic factors with statistically significant odds ratios were the age group of 65 years and over (OR=0.44), persons who identified as Arab/Middle Eastern (OR=0.15), Black (OR=0.17), Indigenous (OR=0.27), Latin American (OR=0.24), South Asian (OR=0.4) and Other Race/Ethnicity (a group those that identified as Southeast Asian, Filipino, West Asian, and Minorities Not Identified Elsewhere, OR=0.17), and those that lived in a household with an income bracket between CAD 25,000-39,999 (OR=0.23).

Regarding the interaction effects, the interaction of the Central Health Region and income bracket between 15000 to 24999 CAD was statistically significant (OR=0.4). In age group and income, significant effects were estimated for those 65 years and older with a household income between CAD 25000-39999 (OR=7.21), and with an income between CAD 40000-59999 (OR=3.12). In the case of race and income, significant effects were estimated for individuals that identified as Arab/Middle Easterners with a family income between 15000 and 24999 (OR=5.68) or income under CAD 15000 (OR=4.21), Black individuals with a family income under CAD 15000 (OR=4.26) or between CAD 15000-24999 (OR=3.35); additionally, there were statistically significant interaction effects in the case of those that identified as belonging to a race not identified elsewhere ("Other") within multiple income brackets: under CAD 15000 (OR=3.73), between 15000 and 24999 (OR 11.2), between 25000 and 39999 CAD (OR=4.67), and with income CAD 40000-59999 (OR=9.58). Finally, there was a significant interaction effect for South Asian individuals with an income under CAD 15000 (OR=3.10).

@@

# create table with the model output
m1 %>%
  tbl_regression(exponentiate = TRUE,
                 label=list(Health_Region ~ "Health Region",
                 Age_group ~ "Age Group")) %>%
    bold_labels()%>%
  as_kable_extra(booktabs = TRUE, longtable=T, linesep="") %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down" argument here.
  kableExtra::kable_styling(latex_options = "repeat_header")
```

>>>>

@tbl-model shows the results of the logistic regression on vaccination status using socio-economic, geographical factors (by Health Region) and their interactions as predictors. The reference groups in each case were set as follows: 16 to 34 years (age group), White Caucasian (Race), Toronto (Health Region), CAD 60,000 and over (Income). Statistically significant odds ratios were identified for individuals with a household income under CAD 25,000 (OR=0.37) and between CAD 25,000-59,999 (OR=0.59), persons who identified as Arab/Middle Eastern (OR=0.31), Black (OR=0.32), Latin American (OR=0.28), Other Race/Ethnicity (a group those that identified as Southeast Asian, Filipino, West Asian, and Minorities Not Identified Elsewhere, OR=0.22), and inviduals living in the West Health Region (OR=1.55).

Moreover, statistically-significant higher odd ratios were determined in the case of income and race; specifically, for individuals with a household income below CAD 25,000 that identified as Arab/Middle Eastern (OR=3.05), Black (3.19), Latin American (OR=2.80), and Other (4.59); and for individuals with a household income between CAD 25,000 and 59,999 that identified as Other racial minority (OR=6.93).

For Health Region and race, significant lower vaccination odds were identified for Black individuals in the Central Health Region (OR=0.44), and Other racial minorities and South Asian individuals that live in the West Health Region (ORs=0.41).

@@

# create table with the model output
m1 %>%
  tbl_regression(exponentiate = TRUE,
                 label=list(Health_Region ~ "Health Region",
                 Age_group_ord ~ "Age Group")) %>%
    bold_labels()%>%
  as_kable_extra(booktabs = TRUE, longtable=T, linesep="") %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down" argument here.
  kableExtra::kable_styling(latex_options = "repeat_header")
```

### 2023-02-21 manuscript/04-discussion.qmd
re-grouped data and ran model, fixed code

Although there was no significant difference between vaccination likelihood among the Health Regions, there was a significant lower vaccination likelihood in the Central Health Region in individuals with a household income between CAD 15,000-24,999 (@tbl-model). This result would be consistent in the context of to Ontario Health, The Central Health Region comprises the former Central, Central West, Mississauga Halton, and North Simcoe Muskoka LHINs with 48.3% immigrant population and 47% of the population identifying as visible minority, the highest percentage among all regions [@ontariohealth].

>>>>

Although there was no significant difference between vaccination odds among the Health Regions, there was a significant lower vaccination odds in the Central Health Region in individuals with a household income between CAD 15,000-24,999 (OR=0.4, @tbl-model). The Central Health Region comprises the former Central, Central West, Mississauga Halton, and North Simcoe Muskoka LHINs with 48.3% of immigrant population and 47% of the population identifying as visible minority, the highest percentage among all Health Regions [@ontariohealth]; moreover, the fact that the Central West LHIN, now within the Central Health Region, had the highest proportion of low-income families in 2018 (12.9\% of males and 16.7\% of females among the lowest 20% of Canadian family incomes [@buajitti2018])

### 2023-02-21 manuscript/appendix.qmd
re-grouped data and ran model, fixed code

It can be seen that there are differences in the distribution of each ethnicity between the Census data and the survey. For example, Arabs/Middle Easteners correspond to 5.9% of the survey responses, but population-wise individuals that identify in this ethnic group correspond to 1.6% of the Ontario population.

>>>>

It can be seen that there are differences in the distribution of each ethnicity between the Census data and the survey. For example, Arabs/Middle Easterner correspond to 5.9% of the survey responses, but population-wise individuals that identify in this ethnic group correspond to 1.6% of the Ontario population.
